---
title: "Hypothesis Testing"
output: html_notebook
---

## What Is It About?

We've already seen a case where testing whether a statement we made was 
plausible or not : while studiying the `juul` dataset we noticed there was a 
difference between the means of `males` and `females`. By looking at the 
confidence interval at 95% of each empirical means we gave an argument 
suggesting there was a significant difference between means of original 
populations and the empirical results were not different because of sampling. 

Hypothesis tests in statistics mainly take the following form : Studying a 
random phenomenon in statistics is generally about studying a sample
singled out from original population. Empirical statistics of sample give 
insight into statistics of whole population when studied statistics is known to 
follow an existing stastical model. This model invovles a parameter which is the 
target of our study. It can take a range of possible values, a subset of which
corresponds to the existant of a specific phenomenon we're investigating. To 
properly formulate hypothesis testing we split the range of possibilities into 
two subsets corresponding either to existance of studied phenomenon or not. 

* The null hypothesis (H_0) reflects to the case when there is no phenomenon
* The alternative hypothesis (H_1) reflects when phenomenon is.

**Here is a standard example.**

You're looking into the prices of personal computers since the eighties, you 
might want to check if average price in comparison to today has changed or not
(you'll need to take inflation into account to make this meaningful). If `m` is 
the empirical expected price of personal computers in the eighties you might 
want to look at the following testing frame:

    * H_0 : today's expected price is `m`
    * H_1 : today's expected price is not `m` anymore.

You could also choose to have the below perspective:

    * H_0 : there is no difference between expected prices in eighties and 
      ninties
    * H_1 : difference is significant.
    
Other testing frames could be about checking for an increase, then for the fact
inflation is not enough to explain increase if any.

As we'll be seeing later, putting down your hypothesis is an essential point in
order to understand what we're looking for, how we're going to check it and 
whether statistical method is robust enough to ensure meaningfulness.

## Student one-sample t-test

This is a standard test, testing whether the mean of a population studied 
through a sequence of mutually indepedent random variables having same normal
law of mean $\mu$ and standard deviation $\sigma$ differs from a given value 
$\mu_0$. Write $\overine{X}_n$ for the sequence means up to rank $n$, $S_n$ for
the one of standard deviation.

* H_0 : mean of population is $\mu_0$
* H_1 : it is not the case.

We've already seen that 
$$ t = \frac{\overline{X}_n - \mu_0}{S_n/\sqrt{n}}$$

should follow a student law having $n-1$ degrees of freedom. In that case this 
random variable has to take values centered and not too far from $0$. Taking 
quantiles $t_{0.025}$ and $t_{0.975}$ for student law, it has $95%$ probability 
of being within $[t_{0.025}, t_{0.975}]$. If it appears not to be so, we are 
inclined to reject the H_0. This is thus done under $t \notin [t_{0.025}, t_{0.975}]$.

To add for this significancy test, statistical tests compute what is called the 
$p$-value. This is the probability of having an extremal value as the one 
previously obtained through the test for $t$. If this probability is higher than
the probability of $t$ falling out of the interval, then we are inclined not to 
reject the null hypothesis. 

Let's try checking the first part of the test by hand.

```{r, example use of t-test for one sample}
n <- 100
dummy <- rnorm(n, mean=8, sd=3)
dummy_bar <- mean(dummy)
dummy_sd <- sd(dummy)

# Null hypothesis is mean equality to 7
t_dummy <- sqrt(n)*(dummy_bar - 7)/dummy_sd
print(t_dummy)
print(qnorm(0.025))
print(qnorm(0.975))
qnorm(0.025) < t_dummy && t_dummy < qnorm(0.975) 
```

We're inclined to reject the null hypothesis. Fortunately there is an R function
that does that : `t.test`. It even gives back a $p$-value that is useful to 
cross-validate our guess. Remember it gives probability of achieving as extreme
as $t$-value we get. If it is higher than the probability of mean being in the 
5% exterior subset, one is advised not to reject null hypothesis.

```{r, using t.test on dummy}
t.test(dummy, mu=7)
```

## Student Two-sample t-test

Consider two sequences of mutually independant random variables having same normal
law each respectively of mean $\mu_1$ and $\mu_2$ and standard deviations
$\sigma_1$ and $\sigma_2$. 

* H_0 : $\sigma_1 = \sigma_2$
* H_1 : $\sigma_1 \neq \sigma_2$

The statistic we compute is the following one (we'll be using indexes to point
out empirical statistics of first and/or second sequence):

$$ t = \frac{\overline{X}_{1, n} - \overline{X}_{2, n}}{\sqrt{(1/n_1)S_{1, n_1}^2 + (1/n_2)S_{2, n_2}^2}}$$

In general this statistic doesn't follow a student law (unless variances of two
groups are the same). It can however be approximated by a law which is a student
one.

```{r, two sample student test}
t.test(juul_female$igf1, juul_male$igf1)
```

R output is to be understood as in the previous case. 

**Previous tests have counterparts called Wilcoxon tests. Can you figure out the**
**difference with the t.tests we've seen here?**

**How would you test variance differences?**

## (Linear) Correlation

Trying to generate random data to use for correlation tests.

```{r, uncorrelated case}
x = rnorm(10000)
y = rnorm(10000)
plot(x, y)
```

```{r}
cor(x, y)
```

```{r, correlated vectors}
u = 3*rnorm(1000, mean=1, sd=5) + 2
v = 2*u + rnorm(1000, sd=3)
plot(u, v)
```

```{r}
cor(u, v)
```

```{r, correlated vectors}
u = 3*rnorm(1000, mean=1, sd=5) + 2
v = -2*u + rnorm(1000, sd=3)
plot(u, v)
```

```{r}
cor(u, v)
cor.test(u, v)
```

```{r, correlated vectors}
u = 3*rnorm(1000, sd=3)
v = u*u + rnorm(1000, sd=8)
plot(u, v)
```

```{r}
cor(u*u, v)
```

```{r}
v = rnorm(1000, sd=0.3)
u = rep(0, 1000) + rnorm(1000, sd=0.01)
plot(v, u)
```

```{r}
cor(v, u)
```

```{r}
cor.test(x, y)
```


