---
title: "Juul Dataset Analysis"
output: html_notebook
---

## Loading Dataset

```{r, Downloading ISwR package}
install.packages("ISwR")
```

```{r, loading package}
library("ISwR")
```

## Digging in

We'll be subsequently using the dataset `juul` which contains medical data 
about growth rate of children. The `igf1` entry is an insulin-like growht factor
and `tanner` is a classification for puberty stages.

```{r, Trying out descriptive statistics on juul}
juul_local <- juul
str(juul_local)
```

```{r}
for (name in c("sex", "menarche", "tanner")){
  juul[[name]] <- factor(juul_local[[name]], ordered=(name == "tanner"))
}
levels(juul_local$sex) <- c("Male", "Female")
levels(juul_local$menarche) <- c("No","Yes")
str(juul_local)
```

```{r}
summary(juul_local)
```

* There are a little more females than males in the dataset;
* Half the sample has age in between 9 and 15. Which suggest rather a young
  population;
* Nearly 25% of the sample didn't get any result for `igf1`, 20% for `tanner`;
* Most `tanner` known results are centered around boundaries.

We aim at undersanding`igf1` variance in term of other features, we'd rather 
eliminate rows of the dataframe corresponding to missing `igf1` values. 

```{r, eliminating missing vales}
juul_local <- juul_local[!is.na(juul_local$igf1), ]
summary(juul_local)
```

Understanding what factors do play role in the `igf1` feature is about linking
it to other features. It would a good point for a starter to look into standard 
statistics per modality of each categorical variable. A good starting point 
can be `sex` feature. Since `menarche` is related to this feature, it can only
be useful in such a subset, adding to that the rare missing values we have for
`sex`.

A way to get what we're looking for is to use the `split` function we've already
seen

```{r, looking into mean per sex}
juul_local <- juul_local[!is.na(juul_local$sex), ]
igf1_sex <- split(juul_local$igf1, juul_local$sex)

statistics_sex <- data.frame(Male=rep(0, 6), Female=rep(0, 6))
rownames(statistics_sex) <- c("min", "Q1", "median", "mean", "Q3", "max")
for (i in names(igf1_sex)){
  group <- igf1_sex[[i]]
  statistics_sex[[i]] <- c(min(group), quantile(group, probs=c(0.25)), median(group), 
                           mean(group), quantile(group, probs=c(0.75)), max(group))
}  

statistics_sex
```

Looking into these results, one might want to say that females `igf1` feature is
higher than males one in average. This is not necessarily the case, and this
impression might be considered as a sampling issue ; you randomly select your 
sample, selection process might give you differences without this being 
significant in the original population you took your sample from. We're 
precisely in the case when one would like to test a statistical hypothesis:
**Are means of both samples significantly different?**. 

```{r, splitting into male and female subsets}
juul_male <- juul_local[juul_local$sex == "Male", ]
juul_female <- juul_local[juul_local$sex == "Female", ]
```

```{r, checking coherence for male dataset}
juul_male$menarche <- NULL
str(juul_male)
```

```{r, checking for female dataset}
str(juul_female)
```

```{r, summary of female dataset}
summary(juul_female)
```

It would be natural looking into the the female dataset to wonder whether 
`menarche` has influence on `igf1`. One would however need to take into account
if this is more relevant than looking into the `tanner` feature.
