---
title: "Pre-Studied Datasets for Final Exams"
output: html_notebook
---

The below code chunks enable you to download four datasets available within the 
R ecosystem. Questions related to these four datasets and an extra one will be
worked out during exam.

You are invited to figure out what the datasets represent and to make a thorough
statistical study of their content.

## The Iris Dataset 

If the `datasets` package wasn't installed on your machine then you'd need to 
install it by executing the following chunk.

```{r, Installing R datasets package, message=FALSE, echo=FALSE}
install.packages('datasets')
```

To load within the R session, you'll need to run the following chunk.

```{r, Loading R datasets package}
library('datasets')
```

This is a standard dataset that have been largely studied during the last 
century. It relates different features of given iris species.

```{r, The iris Dataset}
summary(iris)
```

## The `FARS` Dataset from `gamclass` Package

We'll be first installing the `gamclass` package.

```{r, install package gamclass, message=FALSE, warning=FALSE}
install.packages("gamclass")
```

Loading it next.

```{r, Loading package gamclass}
library('gamclass')
```

The `FARS` dataset is about US fatal car accidents between 1998 and 2010. It has
many features related to each accidents statistics.

```{r, The FARS dataset}
summary(FARS)
```

## The `Baumann` Dataset from the `car` Package

Installing the `car` package first.

```{r, Install car package, message=FALSE, echo=FALSE}
install.packages('car')
```

Loading the `car` package.

```{r, loadint the car package}
library("car")
```

Working out the `Baumann` dataset about *Methods of Teaching Reading Comprehension*. 

```{r, the Baumann dataset}
summary(Baumann)
```

## The `cars` dataset from package `datasets`

This is a dataset relating car speed and stopping distance.

```{r, the cars dataset}
summary(cars)
```

