---
title: "Exam Applied Statistics for Engineers"
output: html_notebook
---

The exam is composed of different case studies where you are expected to show a
reasonable understanding of notions you've seen during the courses. You'll find 
easy up to harder questions in each one of the case studies.

## The Iris Dataset 

In case you haven't installed the datasets package on your machine here is a 
chunck enabling you to do so.

```{r, Installing R datasets package, message=FALSE, echo=FALSE}
install.packages('datasets')
```

1. Start by loading the datasets library and the iris dataset.

2. Check types and first rows of your data.

3. Give a quick summary of the iris dataset's statistical measurements.

4. Only keep subset corresponding to last three features.

5. Single out the the subset of the iris dataset corresponding to setosa specie.
   Be careful to take out colum corresponding to flower type feature.

6. What is the mean of petal length for setosa specie?

7. Compute same mean in the case of the virginica specie?

8. Is the difference in both means significant?

9. Figure out a test that would enable you to single out virginica types from 
   setosa ones. Justify your answer.

10. Assuming petal lengths of virginica and setosa species are normally 
    distributed around their empirical means can you compute the probability
    virginica species would be classified as setosa ones and conversely? 
    (Using your test.)

## The `cars` dataset from package `datasets`

This is a dataset relating car speed and stopping distance.

```{r, the cars dataset}
summary(cars)
```

1. Plot speed against distance.

2. Is there any linear correlation between speed and stopping distance?

3. Can you plot the best linear fit to the previous data?

4. Plot the best linear fit on the scatter plot of given data.

5. Test for quadratic correlations? 

6. Which correlation is more reasonable? Can you figure out a statistical 
   measurement that would account for that?


## The `FARS` Dataset from `gamclass` Package

If you haven't installed the `gamclass` package yet start by running this chunk
of code.

```{r, install package gamclass, message=FALSE, warning=FALSE}
install.packages("gamclass")
```

1. Start loading the `gamclass` package.

2. Drop features corresponding to `caseid`, `airbag`, `restraint`, `inimpact`
   and `D_*`.

3. Recode `sex`, `state`, `modelyr`, `injury` and `year` features into factors. 
   Give "M" and "F" names for `sex` feature level.

4. What are the range of years the satistics have been grouped along? What are 
   the ranges of years' of car model ranging along? Are there any unreasonable 
   values? What can one replace them by? Do so.

5. Plot barplots of accident numbers per year.

6. What would you need to know whether the difference between earliest year and 
   latest year accident count is significant?
   
7. Is there a significant difference between injury 4 levels of earliest year 
   and latest one?

8. Plot barplots of injury level proportions for earliest year and latest one. 
   Are there other significant differences in these proportions?
   
9. Recode the categorical feature `injury` into `< 3` and `> 2` levels. Do the 
   features `airbagDeploy` and `Restraint` affect the recoded injury level? 
